---
title: "Graphics week 10 lab"
author: "Stefany Coxe"
date: "10/31/2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=TRUE}
#install.packages("markdown")
#install.packages("ggplot2")
#install.packages("tidyverse")
#install.packages("Stat2Data")
#install.packages("gapminder")
#install.packages("jtools")
#install.packages("ggrepel")
#install.packages("ggthemes")
#install.packages("broom")
# Install devtools package if necessary
#if(!"devtools" %in% rownames(installed.packages())) install.packages("devtools")
# Install the stable development verions from GitHub
#devtools::install_github("crsh/papaja")
# Install the latest development snapshot from GitHub
#devtools::install_github("crsh/papaja@devel")
#install.packages("kableExtra")
#install.packages("bookdown")
#install.packages("xtable")
#install.package("knitr")
library(markdown)
library(ggplot2)
library(tidyverse)
library(Stat2Data)
library(gapminder)
library(jtools)
library(ggrepel)
library(ggthemes)
library(broom)
library(papaja)
library(kableExtra)
library(bookdown)
library(xtable)
library(knitr)
```

# Adding a specific equation to a plot

We have used `geom_smooth` before to create a general line

- `method = lm` for a straight line

- `method = loess` for a weighted average smooth line

What if you have a more specific equation?

- This example will use a model-based equation

- But you can also specify a certain (e.g., theoretical) equation

## geom_smooth option: add a specific equation to plot

Remember that we had a not-quite-linear relationship between HS GPA and college GPA

```{r}
data(FirstYearGPA)
scatter <- ggplot(data = FirstYearGPA, 
                      aes(x = HSGPA, y = GPA)) +
  geom_point()

scatter + geom_smooth(method = loess, se = FALSE)
```


## Step 1. Run the model

```{r}
cubic_model <- lm(GPA ~ HSGPA + I(HSGPA^2) + I(HSGPA^3), 
                  data = FirstYearGPA)
cubic_model
```

## Step 2. Tidy the model

```{r}
cubic_tidy <- tidy(cubic_model)

print(cubic_tidy)

```

## Step 3 - version 1. Use formula in geom_smooth 

```{r}
cubic_plot1 <- scatter +
  geom_point() +
  geom_smooth(method = lm, 
              formula = y ~ x + I(x^2) + I(x^3), 
              se = FALSE)

cubic_plot1
```

## Step 3 - version 2. Use model values in geom_smooth 

```{r}
cubic_plot2 <- scatter +
  geom_point() +
  geom_smooth(method = nls, 
formula = y ~ intercept + linear*x + 
                quadratic*x*x + cubic*x*x*x, 
se = FALSE,
method.args = list(start = 
c(intercept = cubic_tidy$estimate[1], 
linear = cubic_tidy$estimate[2], 
quadratic = cubic_tidy$estimate[3], 
cubic = cubic_tidy$estimate[4]))) 

cubic_plot2
```

# Two lines with a legend

This is a bit of a cheat to get this to work

You take advantage of the fact that 

- Color arguments generally go *outside* of `aes()` in a geom_smooth

- But if you put the color argument *inside* `aes()`, it treats that as a **label** instead of a color assignment

```{r}
model_colors <- RColorBrewer::brewer.pal(3, "Set1")
cubic_plot3 <- scatter + geom_point() +
  geom_smooth(method = nls, se = FALSE,
formula = y ~ intercept + linear*x + 
  quadratic*x*x + cubic*x*x*x, 
method.args = list(start = 
c(intercept = cubic_tidy$estimate[1], 
linear = cubic_tidy$estimate[2], 
quadratic = cubic_tidy$estimate[3], 
cubic = cubic_tidy$estimate[4])),
aes(color = "Cubic", fill = "Cubic")) +
  geom_smooth(method = lm, aes(color = "Linear", 
                               fill = "Linear"), 
              se = FALSE) +
  scale_color_manual(name = "Models", 
                     values = model_colors) +
  scale_fill_manual(name = "Models", 
                    values = model_colors)

cubic_plot3
```

# Confidence intervals based on your model

Calculate the confidence intervals based on your model with the `predict` function using the optional argument `interval = predict`


```{r}
gpa_cubic_predictions <- predict(object = cubic_model, 
                                 newdata = FirstYearGPA, 
                                 interval = "predict")
head(gpa_cubic_predictions)
gpa_cubic_predict <- cbind(FirstYearGPA, 
                           gpa_cubic_predictions)
```

New variables produced here:

- `fit`: predicted value

- `lwr`: lower CI limit

- `upr`: upper CI limit

```{r}
cubic_plot4 <- scatter + geom_point() +
  geom_line(data = gpa_cubic_predict, 
            aes(x = HSGPA, y = fit)) +
  geom_ribbon(data = gpa_cubic_predict, 
              aes(ymin = lwr, ymax = upr), 
              alpha = .2)

cubic_plot4
```

# nest() and group_by()

We have already used `group_by()` to perform an operation across several groups

- e.g., calculate the mean GDP per capita *for each country*

- One value of mean GDP per capita for each country

The `nest()` function goes one step further and actually splits the dataset up into chunks for each group

- You end up with a **nested table**

- You can run separate analyses and get *separate results* 

```{r}
GPA_nest_genderrace <- FirstYearGPA %>%
  group_by(Male, White) %>%
  nest()

GPA_nest_genderrace
```

## A quick diversion: writing a function

We **use** functions all the time when using R

- `lm()` is a function, `ggplot()` is a function, etc.

We can also **write** functions to do things

- Useful when you'd have to write the same code over and over

- Just call the function instead

```{r}
lm_fxn <- function(dataframe){

lm(GPA ~ HSGPA, data = dataframe)

}
```

This function has one argument, `dataframe`

The function uses the specified dataset to run the regression of HSGPA predicing GPA


## Back to using nest() and group_by()

The `map()` function applies a function to a set of objects

`data` is the name of the column in our nested dataframe that contains the actual data

`lm_fxn` is the function we just created

- This will run the regression specified in `lm_fxn` to each of the nested dataframes in the nested dataset

- We will have 4 separate outputs, one for each group

```{r}
grouped_lm <- GPA_nest_genderrace %>%
  mutate(model = map(data, lm_fxn))

```

Here is the nested dataframe

```{r}
grouped_lm
```

- One column contains the data for each group

- One column contains the regression model for each group


We can apply the `tidy()` function to the nested dataframe

- Again, using the `map()` function

```{r}
tidy_grouped_lm <- grouped_lm %>% 
  mutate(tidied = map(model, tidy))
tidy_grouped_lm
```

Then we can `unnest()` the dataframe to see values

```{r}
out_tidy <- tidy_grouped_lm %>%
  unnest(tidied)
out_tidy
```

## Then filter the dataset

The dataframe includes all the coefficients 

- Here, just intercept and slope

- But say that we really only want to look at slope

```{r}
slopes_only <- out_tidy %>%
  filter(term == "HSGPA")
slopes_only
```

## Then make a plot of the values

```{r}
est_grouped <- ggplot(data = slopes_only,
            mapping = aes(x = as.factor(Male), y = estimate,
                          ymin = estimate - 2*std.error,
                          ymax = estimate + 2*std.error,
                          group = as.factor(White), color = as.factor(White))) +
  geom_errorbar(position = position_dodge(width = 1)) +
  labs(x = "Gender", y = "Estimate", color = "Race")
```

## Then make a plot of the values

```{r}
est_grouped 
```

You could instead do something like use both the intercepts and the slopes to plot simple slopes

- The values are all there, you can do with them what you want

